---
title: "Benchmarking a method against known ground truth"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking a method against known ground truth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

```{r setup}
library(SynthONA)
```

The problem this package addresses is that organisational network methods cannot
normally be validated. Real ONA data is confidential and, more fundamentally, nobody
knows the true answer for it: there is no independent record of who the real brokers
are. A method that reports a plausible-looking set of brokers cannot be shown to be
right or wrong.

Synthetic data solves that, but only if it ships with the answer key and only if it
is degraded the way real data is degraded.

## A specification is an object

```{r}
p <- synthona_params(
  n = 400,
  template = "tech_product",
  topology = "sbm",
  mean_degree = 12,
  within_share = 0.85,
  layers = c("communication", "advice", "trust")
)
p
```

`within_share` is the share of a person's ties expected to stay inside their own
department. It is the dial that controls how siloed the organisation is, and it is
the quantity a community detection method is being asked to recover.

## Generation records what it built

```{r}
d <- synthona_generate(p)
d
d$truth
```

The ground truth is computed on the complete network, before any noise. It is not
re-derived from the observed data later, which is what makes it an answer key rather
than another estimate.

## Recovery on the complete network

With a census of a known network, community detection should do well:

```{r}
g <- synthona_graph(d)
set.seed(47)
louvain <- igraph::cluster_louvain(g)
score_communities(d$truth, louvain)
```

## Recovery under a realistic survey

Real ONA data is collected by survey. Not everyone answers, name generators cap how
many colleagues can be listed, and memory favours strong ties.

```{r}
designs <- list(
  census  = observation_design(),
  good    = observation_design(response_rate = 0.80, name_generator_limit = 10),
  typical = observation_design(response_rate = 0.50, name_generator_limit = 5,
                               recall_probability = 0.80),
  poor    = observation_design(response_rate = 0.30, name_generator_limit = 3,
                               recall_probability = 0.60)
)

results <- do.call(rbind, lapply(names(designs), function(nm) {
  obs <- synthona_observe(d, designs[[nm]])
  gg <- synthona_graph(obs)

  # Louvain is stochastic, so it is seeded to keep this table reproducible.
  # Generation is deterministic; the method under test need not be.
  set.seed(47)
  found <- igraph::cluster_louvain(gg)

  btw <- igraph::betweenness(gg, normalized = TRUE)
  top <- names(sort(btw, decreasing = TRUE))[seq_along(d$truth$brokers)]

  cbind(
    design = nm,
    ties = nrow(obs$edges),
    score_communities(d$truth, found)[, c("ari", "nmi")],
    broker_f1 = score_brokers(d$truth, top)$f1
  )
}))
results
```

The truth object is untouched by observation, so every design is scored against the
same answer key. This is the question the package exists to answer: how much survey
quality does a given method need before its conclusions stop holding?

## Sweeping one parameter

Because the specification is an object, a sweep is an ordinary `lapply`. Tie volume
is held constant, so only the quantity under study varies.

```{r}
sweep <- do.call(rbind, lapply(c(0.40, 0.60, 0.80, 0.95), function(w) {
  ds <- synthona_generate(update_params(p, within_share = w, n = 250))
  obs <- synthona_observe(ds, observation_design(response_rate = 0.6,
                                                 name_generator_limit = 6))
  set.seed(47)
  data.frame(
    within_share = w,
    ari = round(score_communities(
      ds$truth, igraph::cluster_louvain(synthona_graph(obs))
    )$ari, 3)
  )
}))
sweep
```

Recovery improves as the planted structure strengthens, which is the expected
direction. A method that did not show this gradient would be suspect.

## Checking the generator itself

`validate_dataset()` compares realised quantities against the parameters that were
supposed to produce them. Every check can fail.

```{r}
validate_dataset(d)[, c("check_id", "threshold", "measured", "pass")]
```

## Writing datasets out

```{r, eval = FALSE}
synthona_write(d, "output")
read_manifest("output/sbm_n400")$reproduce
```

The manifest records the full specification and the protocol version, so an exported
dataset can be regenerated exactly from its files.
